Speed and strength of an epidemic intervention

### 2020-01-22 doc/speedstrength.tex
Working on language and explanations

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
%% jd: Nice command

@@

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

Much research has prioritized estimates of the basic reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.

Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR_0=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed. 
We argue that the historical primacy of $\RR_0$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.

@@

We model disease incidence using the renewal equation, a simple, flexible framework that can cover a wide range of model structures \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
In our model, disease incidence at time $t$ is given by:
\begin{equation}
i(t) = \int K(\tau, t) i(t-\tau) d\tau.
\end{equation}
Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible, $S(t)$.
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx 1$ \jd{Not necessarily. This is exactly what I meant by my ``basic'' comment.});
under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).

@@

\jd{I would like to hyphenate constant-strength and constant-speed when they are used as adjectives.}

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/constant_intervention.pdf}
\caption{
\textbf{Effects of constant-strength and constant-speed intervention on infection kernels.}
Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and $\RR_0$: 1.5) is shown in black \citep{park2019practical}.
The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
(A) A constant-strength intervention with $\theta = 1.5$ is applied to an Ebola-like infection kernel.
A constant-strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$), the resulting distribution is equivalent to the intrinsic generation-interval distribution ($\hat K(\tau) = g(\tau)$).
(B) A constant-speed intervention with $\phi \approx 0.0267/\mathrm{day}$ is applied to the same kernel.
A constant-speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; when the speed of intervention matches the speed of epidemic ($\phi = r$), the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$). 
}
\figlab{constant}
\end{figure}

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR_0 g(\tau),
	\eqlab{strengthFactors}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution.
The generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR_0$ is thus the integral of $K$.

Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
\begin{equation}
	\hat K(\tau) = (\RR_0/\theta) g(\tau).
\end{equation}
Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR_0$. We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR_0$.

@@

We define the strength of the intervention $L$ to be $\theta = \RR_0/\Rhat$. It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
\begin{equation}
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
	\eqlab{strengthMean}
\end{equation}
In this more general case, we have again that the disease cannot spread when $\theta \geq \RR_0$.

@@

We generalize this idea by allowing the hazard rate $h(\tau)$ at which $K$ is reduced to vary through time, thus:
\begin{equation}
	\hat K(\tau) = K(\tau) \exp\left(-\int_0^\tau h(\sigma) d\sigma\right)
\end{equation}
The associated post-intervention epidemic speed $\hat r$ is given by:
\begin{equation}
	1 = \int \hat K(\tau) \exp(-\hat r\tau) d\tau.	
\end{equation}
We define the speed of a general intervention to be $\phi = r - \hat r$. 
We can then show that $\phi$ is a (sort of) mean satisfying:
\begin{equation}
	1 = \left\langle \frac{\exp(\phi \tau) }{\exp\left(\int_0^\tau h(\sigma) d\sigma\right)} \right\rangle_{b(\tau)}
	\eqlab{speedMean}
\end{equation}
Specifically, the speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.

@@

In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the human immunodeficiency virus (HIV). 
These examples are not meant to provide specific estimates of the effectiveness of intervention strategies; 
instead, we use these examples to demonstrate how strength- and speed-based decompositions can help us better evaluate control strategies under different scenarios.
In particular, we study how the amount of early HIV transmission affects our decision.

@@

As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
Based on the growth rate estimate ($0.452\,\mathrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproductive number is $3.17$.
Increase in the estimate of the amount of early transmission shortens the mean generation time, which in turn decreases the reproductive number (\figref{example}C).
Therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is less early transmission.

>>>>

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\pEarly}{\ensuremath{\tsub{p}{early}}}

@@

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
Epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

 in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}
Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called \Ro, because \RR has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.

Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed. 
We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.

@@

We model disease incidence using the renewal equation, a simple, flexible framework that can cover a wide range of model structures \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
In our model, disease incidence at time $t$ is given by:
\begin{equation}
i(t) = \int K(\tau, t) i(t-\tau) d\tau.
\end{equation}
Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible, $S(t)$.
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx S_0$;
under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/constant_intervention.pdf}
\caption{
\textbf{Effects of constant-strength and constant-speed intervention on infection kernels.}
Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and \Ro: 1.5) is shown in black \citep{park2019practical}.
The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
(A) The effect of a constant-strength intervention with $\theta = 1.5$.
A constant-strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$), the resulting distribution is equivalent to the intrinsic generation-interval distribution ($\hat K(\tau) = g(\tau)$).
(B) A constant-speed intervention with $\phi \approx 0.0267/\mathrm{day}$ is applied to the same kernel.
A constant-speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; when the speed of intervention matches the speed of epidemic ($\phi = r$), the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$). 
}
\figlab{constant}
\end{figure}

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR g(\tau),
	\eqlab{strengthFactors}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution.
The generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR$ is thus the integral of $K$.

Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
\begin{equation}
	\hat K(\tau) = (\RR/\theta) g(\tau).
\end{equation}
Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$. We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR$.

@@

We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$. It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
\begin{equation}
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
	\eqlab{strengthMean}
\end{equation}
In this more general case, we have again that the disease cannot spread when $\theta \geq \RR$.

@@

We generalize this idea by allowing the hazard rate $h(\tau)$ at which $K$ is reduced to vary through time, thus:
\begin{equation}
	\hat K(\tau) = K(\tau) \exp\left(-\int_0^\tau h(\sigma) d\sigma\right)
\end{equation}
The associated post-intervention epidemic speed $\hat r$ is given by:
\begin{equation}
	1 = \int \hat K(\tau) \exp(-\hat r\tau) d\tau.	
\end{equation}
We define the speed of a general intervention to be $\phi = r - \hat r$. 
We can then show that $\phi$ is a (sort of) mean satisfying the implicit equation:
\begin{equation}
	1 = \left\langle \frac{\exp(\phi \tau) }{\exp\left(\int_0^\tau h(\sigma) d\sigma\right)} \right\rangle_{b(\tau)}
	\eqlab{speedMean}
\end{equation}
Specifically, the speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.

@@

In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the human immunodeficiency virus (HIV). 
In particular, we study how the amount of early HIV transmission affects estimates of intervention effectiveness. 
These examples are not detailed estimates for specific scenarios; 
instead, they are meant to demonstrate how strength- and speed-based decompositions can help evaluate control strategies.

@@

The infection kernel is shown in (\figref{example}A) for our baseline value of  
$\tsub{p}{early} = 0.23$.  We assume that the initial speed of the epidemic is $r_0=0.452\,\mathrm{year}^{-1}$; \citep{barron2013eliminating} (see \figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
When transmission is fast, (i.e., when \pEarly\ is large), individuals don't need to transmit as much to achieve this speed, so the estimated value of \Ro\ decreases (\figref{example}C).
Therefore, as \pEarly\ gets smaller, we expect stronger intervention to be required in order to control the disease.
